---
title: "SURE - SUR model estimator example"
author: "4EK608 / 4EK416, Block 3"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lmtest)
library(plm)
library(tidyverse)
library(systemfit)
library(broom)
library(psych)
data <- read.csv("F10_1.csv", sep = ";", header = TRUE)
```

---  

## A Regional Production Model for Public Capital

+ Example amended from Greene, Econometric Analysis (7th ed., 8th ed.).  
+ Based on Munnell (1990) model of productivity of public capital at the state level (USA).  
+ The main equation of the analysis (a Cobb–Douglas production function):  

$$
\log \text{gsp}_{it} = \alpha_i + \beta_{1i} \log \text{pc}_{it} 
+ \beta_{2i} \log \text{hwy}_{it} + \beta_{3i} \log \text{water}_{it} 
+ \beta_{4i} \log \text{util}_{it} + \beta_{5i} \log \text{emp}_{it} 
+ \beta_{6i} \text{unemp}_{it} + \varepsilon_{it}
$$
where  

* gsp      gross state product,  
* pc       private capital,  
* hwy      highway capital,  
* water    water utility capital,  
* util     utility capital,  
* emp      employment (labor),  
* unemp    unemployment rate.  

Annual data for the contiguous 48 states (1970-1986) were combined into 9 regions:  
(this corresponds to the *long panel* format)


* `GF = c("AL", "FL", "LA", "MS")` - Gulf  
* `SW = c("AZ", "NV", "NM", "TX", "UT")`- Southwest    
* `WC = c("CA", "OR", "WA")` - West Coast   
* `MT = c("CO", "ID", "MT", "ND", "SD", "WY")` - Mountain    
* `NE = c("CT", "ME", "MA", "NH", "RI", "VT")` - New England   
* `MA = c("DE", "MD", "NJ", "NY", "PA", "VA")` - Mid Atlantic    
* `SO = c("GA", "NC", "SC", "TN", "WV", "AR")` - South   
* `MW = c("IL", "IN", "KY", "MI", "MN", "OH", "WI")` - Midwest    
* `CN = c("IA", "KS", "MO", "NE", "OK")` - Central   

---  

### Data description & visualization  


Regions

```{r, echo=F, message=F}
# Define regions
GF <- c("AL", "FL", "LA", "MS")
SW <- c("AZ", "NV", "NM", "TX", "UT")
WC <- c("CA", "OR", "WA")
MT <- c("CO", "ID", "MT", "ND", "SD", "WY")
NE <- c("CT", "ME", "MA", "NH", "RI", "VT")
MA <- c("DE", "MD", "NJ", "NY", "PA", "VA")
SO <- c("GA", "NC", "SC", "TN", "WV", "AR")
MW <- c("IL", "IN", "KY", "MI", "MN", "OH", "WI")
CN <- c("IA", "KS", "MO", "NE", "OK")
# Generate region identificator (factor variable)
data$region <- sapply(data$ST_ABB, function(x){
  if(is.element(x, GF)){
    "GF"
  }else if(is.element(x, MW)){
    "MW"
  }else if(is.element(x, MA)){
    "MA"
  }else if(is.element(x, MT)){
    "MT"
  }else if(is.element(x, NE)){
    "NE"
  }else if(is.element(x, SO)){
    "SO"
  }else if(is.element(x, SW)){
    "SW"
  }else if(is.element(x, CN)){
    "CN"
  }else if(is.element(x, WC))
    "WC"
  }
)
data$region <- as.factor(data$region)
# Aggregate data by regions
data_agg <- data %>% 
  group_by(region, YR) %>% 
  summarize(
    gsp = sum(GSP),
    pc = sum(PC),
    hwy = sum(HWY),
    water = sum(WATER),
    util = sum(UTIL),
    emp = sum(EMP),
    unemp = sum(EMP/sum(EMP)*UNEMP)) %>%  # weighted averages used, EMP used as weights
  ungroup()

```

```{r, message=F}
# Aggregated (i.e. into 9 regions) dataset structure:
str(data_agg)

ggplot(data_agg, aes(x=YR,y=log(gsp)))+ 
  geom_point(aes(colour=region))+
  ggtitle("gsp - gross state product")+
  facet_wrap(~region, ncol=3, scales="free")+
  geom_smooth(method= "lm", colour="blue")

ggplot(data_agg, aes(x=unemp,y=gsp))+ 
  geom_point(aes(colour=region))+
  ggtitle("gsp vs unemployment (pairwise, not cet. part. interpretation)")+
  facet_wrap(~region, ncol=3, scales="free")+
  geom_smooth(method= "lm", colour="red")  
```

---  

## Model estimation - *long panels*  
### Alternative specifications & estimators)

##### We start by OLS-estimating 2 different model setups (SURE used in the next step):  

* **Basic pooled model:** we do OLS pooled regression with common coefficients for all equations, information on $\varepsilon_{it}$ contemporaneous covariances ignored). This is just a basic (reference) model.  

* **Equation-by-equation estimates:** individual coefficients estimated by OLS for each equation, information on $\varepsilon_{it}$ contemporaneous covariances ignored).  

##### Next, we assess the efficiency of a SUR/SURE model estimator

* Calculate the variance-covariance matrix $\mathbf{\Sigma}$  
* Based on the results, we evaluate if SUR model FGLS estimation is appropriate (**individual coefficients** for each equation, information on $\varepsilon_{it}$ **contemporaneous covariances used**). 


---  

### OLS estimate with restriction on common coefficients  

**Basic pooled model** (no GLS/FGLS adjustment):  

$$
\begin{bmatrix}
    \mathbf{y}_1 \\ \mathbf{y}_2 \\ \vdots \\ \mathbf{y}_M
    \end{bmatrix} = 
    \begin{bmatrix} 
    \mathbf{X}_1 \\ \mathbf{X}_2 \\ \vdots \\ \mathbf{X}_M 
    \end{bmatrix} \mathbf{\beta} + 
    \begin{bmatrix}
    \mathbf{\varepsilon}_1 \\ \mathbf{\varepsilon}_2 \\ \vdots \\ \mathbf{\varepsilon}_M
    \end{bmatrix} = \mathbf{X\beta}+\mathbf{\varepsilon}
$$
+ where $\mathbf{y}$ is $MT \times 1$ (MT = $9 \times 17 = 153$), $\mathbf{X}$ is $(MT \times K)$ and $\mathbf{\beta}$ is $(K \times 1)$.  
+ OLS estimator: $\mathbf{\hat{\beta}}=(\mathbf{X}^{\prime} \mathbf{X})^{-1} \mathbf{X}^{\prime} \mathbf{y}$  


```{r}
F1 <- log(gsp) ~ log(pc) + log(hwy) + log(water) + log(util) + log(emp) + unemp
#
# As we focus on the long-panel estimation approach,
# we skip the usual TS-related model tests...
OLS1 <- lm(F1,data_agg)
summary(OLS1)
logLik(OLS1)
```

---  

### Equation-by-equation OLS

The stacked-form regression model can be written as:  

$$
\begin{bmatrix}
    \mathbf{y}_1 \\ \mathbf{y}_2 \\ \vdots \\ \mathbf{y}_9
    \end{bmatrix} = 
    \begin{bmatrix} 
    \mathbf{X}_1 & \mathbf{0} & \cdots & \mathbf{0} \\ 
    \mathbf{0} & \mathbf{X}_2 & \cdots & \mathbf{0} \\ 
      &   & \vdots &  \\ 
    \mathbf{0} & \mathbf{0} & \cdots & \mathbf{X}_9 \\ 
    \end{bmatrix}
    \begin{bmatrix}
    \mathbf{\beta}_1 \\ \mathbf{\beta}_2 \\ \vdots \\ \mathbf{\beta}_9
    \end{bmatrix} + 
    \begin{bmatrix}
    \mathbf{\varepsilon}_1 \\ \mathbf{\varepsilon}_2 \\ \vdots \\ \mathbf{\varepsilon}_9
    \end{bmatrix} = \mathbf{X\beta}+\mathbf{\varepsilon}
$$

* Note that $\mathbf{\beta}$ is a $(MK \times 1)$ i.e. $(63 \times 1)$ vector.  
* Separate equation-by-equation OLS is performed,  
* $\mathbf{\hat{\beta}}_i=(\mathbf{X}_i^{\prime} \mathbf{X}_i)^{-1} \mathbf{X}_i^{\prime} \mathbf{y}_i$ for $i=1,2,\dots,9$.  

#### Equation-by-equation OLS estimate (no GLS/FGLS adjustment):  
* possible $\mathbf{\varepsilon}_i$ pairwise covariances (and variance) are ignored in the estimation.  

```{r}
fit <- data_agg %>% 
  group_by(region) %>%
  do(data.frame(tidy(lm(F1, data=.)))) # ?broom:::tidy()
#
as.data.frame(fit[1:63,1:6]) # .. see Table 10.1 (Greene, Econometric analysis, 7,8th ed.)
```

---  

#### Covariance & correlation of OLS residuals from the equation-by-equation OLS:

To see the efficiency of SUR model (FGLS) estimator, we compute contemporaneous covariances among model residuals:

+ $\mathbf{\hat{\Sigma}} = [\hat{\sigma}_{ij}]$  
+ where $\hat{\sigma}_{ij} = cov(\hat{\varepsilon}_{it} \hat{\varepsilon}_{jt})$.  

```{r}
# calculate residuals
resid <- data_agg %>% 
  group_by(region) %>%
  do(model = lm(F1, data=.)) %>%
  mutate(residuals1 = list((model)$residuals) ) %>% # return all resids for each region
  ungroup()
#
# arrange residuals into a "wide-format" dataset (regions in columns)
data_agg$residuals <- unlist(resid$residuals1)
aux.DF <- as.data.frame(data_agg[,c("region","YR","residuals")])
aux.DF2<- tidyr:::spread(aux.DF, key = "region", value = "residuals")
#
# variance-covariance matrix
# .. note the complicated interpretation of covariances....
round(cov(aux.DF2[,-1]),5) # 1st column contains YR...
# For illustration purposes, see also:
# contemporaneous correlations among residuals
# .. can be reconciled to Table 10.2 (Greene, Econometric analysis, 8th ed.)
# .. or to Example 10.1 (Greene, Econometric analysis, 7th ed.)
cor.ci(aux.DF2[,-1])
```


+ There is non-negligible information in the $\mathbf{\Sigma}$ matrix.  
+ Hence SUR model and SURE with GLS/FGLS may provide additional efficiency.  

---  

## SUR/SURE model and its estimators 

```{r}
# re-format data for use in the {systemfit} "SUR" estimator.
# .. currently, this transformation is a necessary step for
# .. proper {systemfit} functioning.
pdata <- pdata.frame(data_agg, c("region","YR"))
class(pdata)
# str(pdata) # uncomemnt to view pdata.frame structure
```


### SUR/SURE estimator (FGLS) of the pooled model  


* Model specification (repeated):

$$
\begin{bmatrix}
    \mathbf{y}_1 \\ \mathbf{y}_2 \\ \vdots \\ \mathbf{y}_M
    \end{bmatrix} = 
    \begin{bmatrix} 
    \mathbf{X}_1 \\ \mathbf{X}_2 \\ \vdots \\ \mathbf{X}_M 
    \end{bmatrix} \mathbf{\beta} + 
    \begin{bmatrix}
    \mathbf{\varepsilon}_1 \\ \mathbf{\varepsilon}_2 \\ \vdots \\ \mathbf{\varepsilon}_M
    \end{bmatrix} = \mathbf{X\beta}+\mathbf{\varepsilon}
$$
* $\mathbf{X}$ is $(MT \times K)$ and $\mathbf{\beta}$ is $(K \times 1)$.

* $\mathbf{\hat{\beta}}_{\text{FGLS}} = [\mathbf{X}^{\prime} (\hat{\mathbf{\Sigma}} \otimes \mathbf{I}_T)^{-1} \mathbf{X}]^{-1} \mathbf{X}^{\prime} (\hat{\mathbf{\Sigma}} \otimes \mathbf{I}_T)^{-1} \mathbf{y}$

---- 

#### SURE (FGLS) of the pooled model: 

```{r}
#
# Model formula (repeated)
# F1 <- log(gsp) ~ log(pc) + log(hwy) + log(water) + log(util) + log(emp) + unemp
#
# Pooled model (common beta vector, FGLS is used)
F10.2.pooled <- systemfit(F1, method = "SUR", data=pdata, pooled=T)
# .. see Tables in Example 10.1 (Greene, Econometric analysis, 7,8th ed.)
#   .. note: with `data=data_agg` /not the panel.data.frame/,
#   .. we get the pooled OLS results
summary(F10.2.pooled, residCov=F) 
```


### FGLS estimator for the general SURE/SUR model  

* Model specification (repeated):  

$$
\begin{bmatrix}
    \mathbf{y}_1 \\ \mathbf{y}_2 \\ \vdots \\ \mathbf{y}_9
    \end{bmatrix} = 
    \begin{bmatrix} 
    \mathbf{X}_1 & \mathbf{0} & \cdots & \mathbf{0} \\ 
    \mathbf{0} & \mathbf{X}_2 & \cdots & \mathbf{0} \\ 
      &   & \vdots &  \\ 
    \mathbf{0} & \mathbf{0} & \cdots & \mathbf{X}_9 \\ 
    \end{bmatrix}
    \begin{bmatrix}
    \mathbf{\beta}_1 \\ \mathbf{\beta}_2 \\ \vdots \\ \mathbf{\beta}_9
    \end{bmatrix} + 
    \begin{bmatrix}
    \mathbf{\varepsilon}_1 \\ \mathbf{\varepsilon}_2 \\ \vdots \\ \mathbf{\varepsilon}_9
    \end{bmatrix} = \mathbf{X\beta}+\mathbf{\varepsilon}
$$

* $\mathbf{X}$ is $(MT \times MK)$ and $\mathbf{\beta}$ is $(MK \times 1)$.

* $\mathbf{\hat{\beta}}_{\text{FGLS}} = [\mathbf{X}^{\prime} (\hat{\mathbf{\Sigma}} \otimes \mathbf{I}_T)^{-1} \mathbf{X}]^{-1} \mathbf{X}^{\prime} (\hat{\mathbf{\Sigma}} \otimes \mathbf{I}_T)^{-1} \mathbf{y}$  

* note that the matrix form of the FGLS remains unchanged for both $\mathbf{X}$ setups.  

--- 

#### SURE (FGLS) of the general long-panel model: 

```{r}
#
# (individual beta vectors for each equation, FGLS)
F10.1.SUR <- systemfit(F1, method = "SUR", data=pdata)
summary(F10.1.SUR, residCov=F) # .. see Example 10.1 (Greene, Econometric analysis, 7,8th ed.)
```

---  

#### Finally, we compare the "pooled" and general SUR/SURE models  
(i.e. we test the pooling restrictions):

```{r}
#
# LR test for common beta vector restriction:
lrtest(F10.2.pooled,F10.1.SUR)
```

--- 

##  Supervised work: apply SUR model estimator to the "Grunfeld" dataset

```{r}
data("Grunfeld", package="plm")
# ?Grunfeld
class(Grunfeld) 
str(Grunfeld) 
```

#### Supervised work tasks:

1. Convert `Grunfeld` to `pdata.frame` format.  

2. Estimate `inv ~ value + capital` using a pooled OLS approach.  

3. Perform equation-by-equation OLS estimation for each of the 10 companies.  
4. Assess contemporaneous covariance/correlation of residuals from step (3).  
4. Estimate the pooled SUR model with restricted (identical) beta parameters, use FGLS.  

5. Estimate a general-form SUR model, use FGLS.  

6. Evaluate the validity of restrictions placed in step (4).  


--- 






