---
title: "Partial/Marginal effects and their standard errors"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(knitr)
require(ggplot2)
require(lmtest)
require(sandwich)
require(margins)
require(car)
require(RcmdrMisc)
```


### Data: 
```{r}
attend <- load("attend.RData")
print(desc)
m1 <- mean(data$atndrte)
m2 <- mean(data$priGPA)
m1 # mean(data$atndrte)
m2 # mean(data$priGPA)
``` 
 

***

# Supervised work:

#### 1) We expand the `lm2` model to include $\textit{priGPA}^2$ and $\textit{ACT}^2$ as regressors:

$$\textit{stndfnl}_{\,i} = \beta_0 + \beta_1 \textit{atndrte}_i
+ \beta_2 \textit{priGPA}_i + \beta_3 \textit{ACT}_i  
+ \beta_4 \textit{priGPA}_i^2 + \beta_5 \textit{ACT}_i^2 
+ \beta_6 (\textit{atndrte}_i \times \textit{priGPA}_i  ) 
+ u_i$$

##### Estimate the model

```{r}
# Fill-in and uncomment the following lines
lm4<-lm(stndfnl ~ atndrte + priGPA + ACT +I(priGPA^2) + I(ACT^2) + atndrte:priGPA, data=data) # repeated
summary(lm4)
```

##### Test $H_0: \beta_1 = \beta_6 = 0$
```{r}
# Fill-in and uncomment the following lines
# ?linearHypothesis
linearHypothesis(lm4, c("atndrte = 0", "atndrte:priGPA =0") )
```

For results verification, see Wooldridge, Introductory econometrics, Example 6.3


#### 2) Use R (`DeltaMethod`) to calculate marginal effects and standard errors. Use mean values of all regressors as your "interesting" observations.

```{r}
# 
m3 <- mean(data$ACT)
# 
# Effect and s.e. of atndrte
DeltaMethod(lm4,"(b1+b6*m2)")
#
# Effect and s.e. of priGPA
DeltaMethod(lm4,"(b2+2*b4*m2+b6*m1)")
#
# Effect and s.e. of ACT
DeltaMethod(lm4,"(b3+2*b5*m3)")
#
```

#### 3) Use R (`margins`) to calculate APEs of all regressors, interpret


```{r}
# APEs 
summary(margins(lm4))
```

***


For additional examples, see `margins`: [https://cran.r-project.org/web/packages/margins/vignettes/Introduction.html](https://cran.r-project.org/web/packages/margins/vignettes/Introduction.html).

