---
title: "Monte Carlo simulation example"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lattice)
library(lmtest)

```

***

# Introduction

**Monte Carlo simulation example**

In this example, we compare the power of two tests for autocorrelation: 

* Durbin-Watson 
* Breusch-Godfrey

We use two different specifications of a simulated linear regression for this purpose:

$$ y_t = \beta_0 + \beta_1 t + u_t $$

and

$$ y_t = \beta_0 + \beta_1 y_{t-1} + u_t $$
where the regression coefficients of the simulated model are: $\beta_0 = 0.25$ and $\beta_1 = -0.75$ for both our regression specifications (deterministic trend, dynamic model). $u_t$ is assumed stationary, with mean zero. 

We want to evaluate the small-sample powers of the two tests ($\alpha = 0.05$) under the following scenarios:

*  Autocorrelation: $\rho = 0, 0.2, 0.4, 0.6, 0.8, 0.9, 0.95, 0.99$
*  Sample size: $n = 15, 30, 50$.



### Data: 

For this Monte Carlo example, we need to define the DGP process first.

```{r, echo=T}
dgp <- function(nobs = 15, model = c("trend", "dynamic"),
                corr = 0, coef = c(0.25, -0.75), sd = 1)
{
model <- match.arg(model)
coef <- rep(coef, length.out = 2)
err <- as.vector(filter(rnorm(nobs, sd = sd), corr, method = "recursive"))
if(model == "trend") {
   x <- 1:nobs
   y <- coef[1] + coef[2] * x + err
} else {
   y <- rep(NA, nobs)
   y[1] <- coef[1] + err[1]
   for(i in 2:nobs)
     y[i] <- coef[1] + coef[2] * y[i-1] + err[i]
     x <- c(0, y[1:(nobs-1)])
}
return(data.frame(y = y, x = x))
}
``` 

### Basic data plot:

(one example of simulated data, $n=50$)

```{r,echo=T}
df1 <- dgp(model="trend", nobs = 50)
plot(df1$y~df1$x,main="Trend")
df2 <- dgp(model="dynamic", nobs = 50)
plot(df2$y~df2$x,main="Dynamic")
```

***

### Next, we define a function that evaluates test powers:


```{r}
simpower <- function(nrep = 100, size = 0.05, ...)
{
pval <- matrix(rep(NA, 2 * nrep), ncol = 2)
colnames(pval) <- c("dwtest", "bgtest")
for(i in 1:nrep) {
  dat <- dgp(...)
  pval[i,1] <- dwtest(y ~ x, data = dat,
  alternative = "two.sided")$p.value
  pval[i,2] <- bgtest(y ~ x, data = dat)$p.value
}
return(colMeans(pval < size))
}                      
```

Higher precision of this function may be obtained by increasing the `nrep` argument. However, computation time may be a prohibitive factor here.


***

#### We prepare a function that does the actual Monte Carlo simulation task:
```{r}
simulation <- function(corr = c(0, 0.2, 0.4, 0.6, 0.8, 0.9, 0.95, 0.99), 
                       nobs = c(15, 30, 50), 
                       model = c("trend", "dynamic"), ...)
{
prs <- expand.grid(corr = corr, nobs = nobs, model = model)
nprs <- nrow(prs)
pow <- matrix(rep(NA, 2 * nprs), ncol = 2)
for(i in 1:nprs) pow[i,] <- simpower(corr = prs[i,1],
   nobs = prs[i,2], model = as.character(prs[i,3]), ...)
rval <- rbind(prs, prs)
rval$test <- factor(rep(1:2, c(nprs, nprs)),
labels = c("dwtest", "bgtest"))
rval$power <- c(pow[,1], pow[,2])
rval$nobs <- factor(rval$nobs)
return(rval)
}
```

***

#### Now, we can run the simulation:

```{r}
set.seed(987)
psim <- simulation()
str(psim)
```

***

#### Simulation output can be summarized and plotted:

Test powers, i.e. $H_0$ rejection probabilities for increasing $\rho$ can be observed from the table and from plots:

```{r}
tab <- xtabs(power ~ corr + test + model + nobs, data = psim)
ftable(tab, row.vars = c("model", "nobs", "test"), col.vars = "corr")
xyplot(power ~ corr | model + nobs, groups = ~ test, data = psim, type = "b")
```


* Durbin-Watson test (blue) performs better in the trend model - the difference is only pronounced for smaller samples and lower $\rho$. 
* In the dynamic model, Durbin-Watson test has almost no power except for
very high correlations. 
* Breusch-Godfrey test (pink) performs reasonably well, unless $n$ is very small.





*** 

* Example amended from the textbook Gentleman, Hornik, Parmigiani: Use R!
* For detailed discussion see chapter 7 of [Use R!](https://link.springer.com/content/pdf/bfm%3A978-0-387-77318-6%2F1.pdf). 
